Inferring community characteristics in labelled networks

### 2021-05-20 nips/experiments.tex
More work on experiments

We apply the developed methods to a variety of datasets, to show the power of the method across different scenarios. We choose datasets to span a range of node counts $N$, edge counts $E$ and feature space dimension $D$.

@@

We require metrics to assess the relative model evidence for the FFBM. This can be split into two separate components: the microcanonical SBM fit (concerned with the $b$-samples) and the fit of the feature-to-block generator (concerned with the $\theta$-samples). For each of these we can evaluate the mean of the unnormalised log-target of each Markov chain.
%
\begin{equation}
	\bar{S}_b \coloneqq \frac{1}{T_b} \sum_{t=1}^{T_b} S \left( b^{(t)} \right) \qquad \textrm{and} \qquad
	\bar{U}_\theta \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} U \left( \theta^{(t)} \right)
\end{equation}
%
The lower these quantities, the better the fit of each stage of the model. To allow for rough comparison between datasets, we divide these quantities by the number of vertices in the graph $N$. Table \ref{tab:results} summarises the results for each experiment.

\begin{table}[!h]
	\centering
	\caption{FFBM fit for the datasets}
	\label{tab:results}
	\begin{tabular}{c|ccc|c|cc}
		Dataset & $N$ & $E$ & $D$ & $B$ & $\bar{S}_b /N$ & $\bar{U}_\theta /N$ \\ \hline
		Political books & 105  & 441 & 3 & 3 & 11.7 & 0.707 \\
		Primary school & 238 & 5539 & 13 &  10 & 45.9  & 1.61 \\
		FB egonet & 747 & 30025 & 480 & 10  & 66.9 & 7.47
	\end{tabular}
\end{table}

>>>>

We apply the developed methods to a variety of datasets, to show the power of the method across different scenarios. We choose datasets to span a range of node counts $N$, edge counts $E$ and feature space dimension $D$. We consider the following datasets:

@@

We require metrics to assess the relative model evidence for the FFBM. This can be split into two separate components: the microcanonical SBM fit (concerned with the $b$-samples) and the fit of the feature-to-block generator (concerned with the $\theta$-samples). The SBM has been extensively analysed in other works \cite{Peixoto-Bayesian-Microcanonical} so that will not be the emphasis of our analysis. We use a simple metric to gauge its fit: the mean description length per node of the $b$-samples (equation \ref{eqn:mean-dl}).
%
\begin{equation}
	\frac{1}{N} \bar{S}_b \coloneqq \frac{1}{N |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right)
	\label{eqn:mean-dl}
\end{equation}
%
However, to assess the performance of the feature-to-block predictor, we must partition the vertex set $[N]$ into training and test sets -- $\Gcal_0$ and $\Gcal_1$ respectively\footnote{We choose to randomly create the partition on each experiment run such that 80\% of the available vertices form our training set $\Gcal$ and the remaining 20\% are held out to form our test set $\Gcal^C$}. The $b$-chain is run using the whole network but we only use vertices $v \in \Gcal$ to train the $\theta$-chain. As $|\Gcal_0| \neq |\Gcal_1|$ sets, we cannot use the un-normalised log target $U$ (equation \ref{eqn:U-form}) for comparison between the two, as the prior term will be equal between the two but the total cross-entropy loss will be scaled by the size of each set. We therefore must use the average cross-entropy loss over each set (equation \ref{eqn:cross-entropy-loss}).
%
\begin{equation}
	\bar{\Lcal}_a \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_a \left( \theta^{(t)} \right)
	\qquad \textrm{where} \qquad
	\Lcal_a \left( \theta^{(t)} \right) \coloneqq \frac{1}{|\Gcal_a|} \sum_{i \in \Gcal_a}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)}
	\label{eqn:cross-entropy-loss}
\end{equation}
%
Where $a \in \{0, 1\}$ has been introduced to toggle between training and test sets. Table \ref{tab:results} summarises the results for each experiment. We run each experiment 5 times with the same hyperparameter values\footnote{TODO hyperparam values} and report the mean plus or minus one standard deviation.

However, we also wish to determine the power of the dimensionality reduction method. For this we discard all features that satisfy equation \ref{eqn:discard}, to reduce the dimension from $D$ to $D'$. We then retrain the feature-block predictor using just the retained features and report the loss over the training and test sets for the reduced classifier -- denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. These are also

\begin{table}[!h]
	\centering
	\caption{Table of results}
	\label{tab:results}
	\begin{tabular}{c|ccc|c|c|cc|c|cc}
		Dataset & $N$ & $E$ & $D$ & $B$ & $\bar{S}_b /N$ & $\bar{\Lcal}_0$ & $\bar{\Lcal}_1$ & $D'$ & $\bar{\Lcal}_0'$ & $\bar{\Lcal}_1'$  \\ \hline
		Polbooks & 105  & 441 & 3 & 3 & 11.7 & &&--&--&-- \\
		School & 238 & 5539 & 13 &  10 & 45.9  & &&10&& \\
		FB egonet & 747 & 30025 & 480 & 10  & 66.9 & &&&&
	\end{tabular}
\end{table}

### 2021-05-20 nips/inference.tex
More work on experiments

The invariant distribution we wish to target for the $\theta$ samples is the posterior of $\theta$ given the values of the pair $(X, b)$. We write this as follows:
%
\begin{align}
	p(\theta | X, b) &\propto p(b | X, \theta) p(\theta) = \pi_\theta (\theta) \propto  \exp \left( - U(\theta) \right) \\
	\therefore U(\theta) &= - \left( \log p(b | X, \theta) + \log p(\theta) \right) + \textrm{const}
\end{align}
%
Where we have introduced $U(\theta)$ equal to the negative log posterior. This $U(\theta)$ term makes subsequent analysis more concise. Each of the constituent terms of $U(\theta)$ is easily computed (equation \ref{eqn:U-constituent-terms}). To simplify notation, we define $y_{ij} \coloneqq \one \left\{ b_i = j \right\}$ and $a_{ij} \coloneqq \phi_j(x_i; \theta)$.
%
\begin{equation}
	\log p(b | X, \theta) = \sum_{i=1}^{N} \sum_{j=1}^{B} y_{ij} \log a_{ij}  \quad \textrm{and} \quad
	\log p(\theta) = -\frac{(D+1)(B)}{2} \log 2\pi - \frac{1}{2 \sigma_\theta^2} || \theta ||^2
	\label{eqn:U-constituent-terms}
\end{equation}
%
This means that if we ignore constant terms we can write $U(\theta)$ as in equation \ref{eqn:U-form}. We see that the prior effectively introduces a regularisation term. Note that $||\theta||^2 = \sum_{i} \theta_{i}^2 = \sum_{j=1}^{B} ||w_j||^2$ is the Euclidean norm of the vector of parameters $\theta$.
%
\begin{equation}
	U(\theta) = \left( \sum_{i=1}^{N} \sum_{j=1}^{B} y_{ij} \log \frac{1}{a_{ij}} \right)
	+ \frac{1}{2\sigma_\theta^2} ||\theta||^2
	\label{eqn:U-form}
\end{equation}
%
$U(\theta)$ in equation \ref{eqn:U-form} appears a typical objective function to be minimised for neural network training. The first term is the cross-entropy between actual and predicted labels. The second term - introduced by the prior - brings a form of regularisation, preventing over-fitting. In traditional applications we only seek the value of $\theta$ that minimises the objective function $U(\theta)$, which in our case would yield the maximum a posteriori (MAP) estimate. This is often done through some kind of gradient descent as $\nabla U$ is easily computable (equation \ref{eqn:U-derivative}).

However, our goal is not to find the MAP estimate but to draw samples from the posterior $\pi_\theta(\cdot) \propto \exp(-U(\cdot))$. As discussed earlier, given the invariant $\pi_\theta(\cdot)$, it is sufficient to specify a proposal distribution and then apply a MH accept-reject step to ensure detailed balance of the Markov Chain. Nevertheless, we can use $\nabla U$ as a useful heuristic to bias our proposal towards regions of higher target density. We therefore adopt the Metropolis Adjusted Langevin Algorithm (MALA) - first proposed by \citet{mala-tweedie} - which leverages just that. Given the current sample $\theta$, we propose a new sample $\theta'$ according to equation \ref{eqn:theta-update}
%
\begin{equation}
	\theta' = \theta - h \nabla U(\theta) + \sqrt{2h} \cdot \xi
	\label{eqn:theta-update}
\end{equation}
%
Where $\xi \sim \Gaussian(0, I)$ and $h$ is a step-size parameter - which may vary with the sample index. Indeed without the injected noise term, this is equivalent to gradient descent. We require the noise term to fully explore the parameter space. As such the proposal distribution, is a simple multivariate Gaussian which can be easily evaluated.
%
\begin{equation}
	q_\theta(\theta, \theta') = \Gaussian \left( \theta' ; \theta - h \nabla U(\theta), 2h I \right)
\end{equation}
%
The term $\nabla U$ has an easy to compute analytic form (derived in Appendix \ref{appdx:gradu}). By noting that $\theta = \{w_k\}_{k=1}^{B}$, we write the derivative with respect to each $w_k$ as:
%
\begin{equation}
	\frac{\partial U}{\partial w_k} = - \left( \sum_{i=1}^{N} \Big\{ \tilde{x}_i (y_{ik} - a_{ik}) \Big\} - \frac{w_k}{\sigma_\theta^2} \right)
	\label{eqn:U-derivative}
\end{equation}
%
After a proposed move is generated, in typical Metropolis-Hastings fashion we accept the move with probability $\alpha_\theta$.
%
\begin{equation}
	\alpha_\theta(\theta, \theta') = \min \left( 
	\exp \left( U(\theta) - U(\theta')\right)
	\frac{ 
		q_\theta(\theta', \theta)
	}{
		q_\theta(\theta, \theta')
	} 
	, 1 \right)
\end{equation}
%
This fully specifies, the sampling procedure to generate $\{\theta^{(t)}\}_{t=1}^T$. Bear in mind that each $\theta^{(t)}$ update step uses its corresponding $b^{(t)}$ block membership sample.

@@

@@

Once we have the samples $\left\{ \theta^{(t)} \right\}_{t=1}^{T_\theta} \sim p(\theta | A, X)$, we can compute the empirical mean and standard deviation of each component of $\theta$. Switching back to matrix notation we define $\theta = W$, such that $W_{ij}$ is the weight component for block $i$ and feature $j$, we can define:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma_{ij}} \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2
\end{equation}
%
A simple heuristic to discard the least important features requires specifying a cutoff $c > 0$ and a multiplier $k > 0$. We then discard all features $j$ wuch that equation \ref{eqn:discard} is satisfied.
%
\begin{equation}
	(\hat{\mu}_{ij} - k \hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij}) \cap (-c, +c) = \emptyset \quad \forall i \in \{1,2 \dots B\}
	\label{eqn:discard}
\end{equation}
%
Intuitively, this means discarding any feature for which $\hat{\mu}_{ij} \pm k\hat{\sigma_{ij}}$ lies within or spans the null region $(-c, c)$. If we were to use the Laplace approximation for the posterior $p(W_{ij} | A, X) \approx \Gaussian(W_{ij}; \mu_{ij}, \sigma_{ij})$, then this is effectively a hypothesis test on the value of $W_{ij}$.
%
\begin{equation}
	\begin{aligned}
		H_0&: |W_{ij}| \leq c \\
		H_1&: |W_{ij}| > c
	\end{aligned}
\end{equation}
%
With the multiplier $k$ specifying the degree of significance of the result (equivalent to the probability of a type-1 error). However, as the approximation is not exact we will only treat this dimensionality reduction method as a useful heuristic and not an exact method.

>>>>

The invariant distribution we wish to target for the $\theta$ samples is the posterior of $\theta$ given the values of the pair $(X, b)$. We write this as follows:
%
\begin{align}
	p(\theta | X, b) &\propto p(b | X, \theta) p(\theta) = \pi_\theta (\theta) \propto  \exp \left( - U(\theta) \right) \\
	\therefore U(\theta) &= - \left( \log p(b | X, \theta) + \log p(\theta) \right) + \textrm{const}
\end{align}
%
Where we have introduced $U(\theta)$ equal to the negative log posterior. This $U(\theta)$ term makes subsequent analysis more concise. Each of the constituent terms of $U(\theta)$ is easily computed (equation \ref{eqn:U-constituent-terms}). To simplify notation, we define $y_{ij} \coloneqq \one \left\{ b_i = j \right\}$ and $a_{ij} \coloneqq \phi_j(x_i; \theta)$.
%
\begin{equation}
	\log p(b | X, \theta) = \sum_{i=1}^{N} \sum_{j=1}^{B} y_{ij} \log a_{ij}  \quad \textrm{and} \quad
	\log p(\theta) = -\frac{(D+1)(B)}{2} \log 2\pi - \frac{1}{2 \sigma_\theta^2} || \theta ||^2
	\label{eqn:U-constituent-terms}
\end{equation}
%
This means that if we ignore constant terms we can write $U(\theta)$ as in equation \ref{eqn:U-form}. We see that the prior effectively introduces a regularisation term. Note that $||\theta||^2 = \sum_{i} \theta_{i}^2 = \sum_{j=1}^{B} ||w_j||^2$ is the Euclidean norm of the vector of parameters $\theta$.
%
\begin{equation}
	U(\theta) = \left( \sum_{i=1}^{N} \sum_{j=1}^{B} y_{ij} \log \frac{1}{a_{ij}} \right)
	+ \frac{1}{2\sigma_\theta^2} ||\theta||^2 = N \cdot \Lcal(\theta) + \frac{1}{2\sigma_\theta^2} ||\theta||^2
	\label{eqn:U-form}
\end{equation}
%
$U(\theta)$ in equation \ref{eqn:U-form} appears a typical objective function to be minimised for neural network training. The first term -- which we collect into $N \cdot \Lcal(\theta)$ -- is the cross-entropy between graph-predicted and feature-predicted block memberships. We denote the average cross-entropy $\Lcal(\theta)$ as it will become useful later on when performing experiments. The second term of equation \ref{eqn:U-form} -- introduced by the prior -- brings a form of regularisation, preventing over-fitting. In traditional applications we only seek the value of $\theta$ that minimises the objective function $U(\theta)$, which in our case would yield the maximum a posteriori (MAP) estimate. This is often done through some kind of gradient descent as $\nabla U$ is easily computable (equation \ref{eqn:U-derivative}).

However, our goal is not to find the MAP estimate but to draw samples from the posterior $\pi_\theta(\cdot) \propto \exp(-U(\cdot))$. As discussed earlier, given the invariant $\pi_\theta(\cdot)$, it is sufficient to specify a proposal distribution and then apply a MH accept-reject step to ensure detailed balance of the Markov Chain. Nevertheless, we can use $\nabla U$ as a useful heuristic to bias our proposal towards regions of higher target density. We therefore adopt the Metropolis Adjusted Langevin Algorithm (MALA) -- first proposed by \citet{mala-tweedie} -- which leverages just that. Given the current sample $\theta$, we propose a new sample $\theta'$ according to equation \ref{eqn:theta-update}
%
\begin{equation}
	\theta' = \theta - h \nabla U(\theta) + \sqrt{2h} \cdot \xi
	\label{eqn:theta-update}
\end{equation}
%
Where $\xi \sim \Gaussian(0, I)$ and $h$ is a step-size parameter - which may vary with the sample index. Indeed without the injected noise term, this is equivalent to gradient descent. We require the noise term to fully explore the parameter space. As such the proposal distribution, is a simple multivariate Gaussian which can be easily evaluated.
%
\begin{equation}
	q_\theta(\theta, \theta') = \Gaussian \left( \theta' ; \theta - h \nabla U(\theta), 2h I \right)
\end{equation}
%
The term $\nabla U$ has an easy to compute analytic form (derived in Appendix \ref{appdx:gradu}). By noting that $\theta = \{w_k\}_{k=1}^{B}$, we write the derivative with respect to each $w_k$ as:
%
\begin{equation}
	\frac{\partial U}{\partial w_k} = - \left( \sum_{i=1}^{N} \Big\{ \tilde{x}_i (y_{ik} - a_{ik}) \Big\} - \frac{w_k}{\sigma_\theta^2} \right)
	\label{eqn:U-derivative}
\end{equation}
%
After a proposed move is generated, in typical Metropolis-Hastings fashion we accept the move with probability $\alpha_\theta$.
%
\begin{equation}
	\alpha_\theta(\theta, \theta') = \min \left( 
	\exp \left( U(\theta) - U(\theta')\right)
	\frac{ 
		q_\theta(\theta', \theta)
	}{
		q_\theta(\theta, \theta')
	} 
	, 1 \right)
\end{equation}
%
This fully specifies, the sampling procedure to generate $\{\theta^{(t)}\}_{t=1}^T$. Bear in mind that each $\theta^{(t)}$ update step uses its corresponding $b^{(t)}$ block membership sample.

@@

TODO: burn-in and thinning

@@

Once we have the samples $\left\{ \theta^{(t)} \right\}_{t=1}^{T_\theta} \sim p(\theta | A, X)$, we can compute the empirical mean and standard deviation of each component of $\theta$. Switching back to matrix notation we define $\theta = W$, such that $W_{ij}$ is the weight component for block $i$ and feature $j$, we can define:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma}_{ij} \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2
\end{equation}
%
A simple heuristic to discard the least important features requires specifying a cutoff $c > 0$ and a multiplier $k > 0$. We then discard all features $j$ such that equation \ref{eqn:discard} is satisfied.
%
\begin{equation}
	(\hat{\mu}_{ij} - k \hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij}) \cap (-c, +c) = \emptyset \quad \forall i \in \{1,2 \dots B\}
	\label{eqn:discard}
\end{equation}
%
Intuitively, this means discarding any feature for which $\hat{\mu}_{ij} \pm k\hat{\sigma_{ij}}$ lies within or spans the null region $(-c, c)$. If we were to use the Laplace approximation for the posterior $p(W_{ij} | A, X) \approx \Gaussian(W_{ij}; \mu_{ij}, \sigma_{ij})$, then this is effectively a hypothesis test on the value of $W_{ij}$.
%
\begin{equation}
	\begin{aligned}
		H_0&: |W_{ij}| \leq c \\
		H_1&: |W_{ij}| > c
	\end{aligned}
\end{equation}
%
With the multiplier $k$ specifying the degree of significance of the result (equivalent to the probability of a type-1 error). However, as the approximation is not exact we will only treat this dimensionality reduction method as a useful heuristic and not an exact method.
